Efficient Handling of String-Number Conversion

### 2020-03-30 final-version/main.tex
fixing the comments of reviewers A & D
deleting files

\usepackage{flushend} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\hide}[1]{}
\newcommand{\tool}{{\textsf{Z3-Trau}}}
\newcommand{\dr}{\mathcal{R}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\integers}{\mathbb{Z}}
\newcommand{\todo}[1]{{\color{blue}TODO: #1}}
\newcommand{\lh}[1]{{\color{orange}Lukas: #1}}
\newcommand{\changed}[1]{{\color{blue}#1}}

@@

However, the formula is satisfiable when $x=``00010"$. Observe that $\sti{``00010"}=10$. The key is that even for a bounded integer, the corresponding numeral can be of an unbounded length with arbitrarily many trailing `$0$'s at the front. All numbers with up to $m$-digits can be however still handled without having to solve exponential constraints. It is enough to equip the initial state of the PFA with a  $0$-self-loop.

@@

	\caption{PFA for string-number conversion constraints}
	\label{fig:sfa_its}
\end{figure}

%Such PFA handles bounded integers (e.g., 10-digits or 64-bits) precisely and under-approximate the possible values of unbounded integers. It works very efficiently because we just need to use linear constraints to characterize the relation between $x$ and $n$. 
Numeric PFA with these restrictions would satisfy our primary objective, that is, they would induce linear formulae and would ``easily" and completely cover all numerals.
%
A last problem still needs to be solved before they can be efficient in practice. 
%
Recall that the character variables can be assigned $\epsilon$.  
%
Therefore, a single chain of $k$ interesting digits, $k\leq m$, can be by $A^m$ represented in $k\choose m$ ways, each corresponding to one possible interleaving of $k$ digits with $m-k$ epsilons. 
%
%\lh{I am not sure whether it is really impossible to make a small formula}
%
This may  lead to a formula of   exponential size  when defining  the value of $n$.
%
In order to eliminate this potential blow-up in the size of the formula, we add to $\paf^m$ an additional constraint that forces all epsilons to be \emph{shifted} behind the least significant digit. This will leave us with only one interleaving. This is the formula
$$
\Psi_{\mathsf{shift}}^m \defeq \bigwedge_{1\leq i \leq m} v_i \neq \noepsilon \implies v_{i-1}\neq \noepsilon \ .
$$
Last, since this restriction is meaningful only when the string is indeed a numeral,  
we also define the constraint representing the strings which are not numerals, the formula
$$
\Psi_{\mathsf{NaN}}^m \defeq  \bigvee_{i\in [1,m]} v_i>9
$$ 
and define the final form of the interpretation restriction used by $A^m$ as 
$$
\paf^m \defeq  \Psi_{\mathsf{NaN}}^m \lor (\Psi_{v_0} \land \Psi_{\mathsf{shift}}^m)\ .
$$
Consequently, we design our domain restrictions $\dr$ so that for string variables $x$ that appear within string-integer constraints,  $\dr(x)$ is a number PFA $(A^m,\paf^m),m\in \nat$.

@@

%
%
%$$
%\mathsf{toNumber(k)}\defeq (n = v_1\times 10^{k-1} + v_2\times 10^{k-2} + \ldots + v_{k})
%$$
%
%
%
%
%$$\Psi_{\mathsf{NaN}} \defeq  \bigvee_{i\in [1,m]} v_i>9$$
%where $ $
%To explain this encoding, we first define two macro function:
%
%$$\begin{array}{rl}
%\mathsf{epsilonAfter(k)}\defeq& (0\leq v_1,\ldots, v_k\leq 9) \wedge\\
%&v_{k+1}=\ldots =v_m = -1\\
%\mathsf{toNumber(k)}\defeq& (n = v_1\times 10^{k-1} + v_2\times 10^{k-2} + \ldots + v_{k})
%\end{array}
%$$

%The function $\mathsf{epsilonAfter(k)}$ forces the first $k$ variables to be assigned some number and those after $k$ be assigned $\epsilon$. Recall that we use $-1$ to represent $\epsilon$. The function $\mathsf{toNumber(k)}$ converts the first $k$ variables to the corresponding number in $n$. \yfc{say shifting $\epsilon$ to the right will not change the language}

%The condition that when $x$ is ``not a number" is specified by the following formula.
%
%$$
%\Psi_{\mathit{NaN}} \defeq  \bigvee_{i\in [1,m]} v_i>9$$
%
%Then we can specify the relation between $n$ and the character variables as follows:
%$$\Psi_n\defeq  (n=-1 \wedge \Psi_{\mathit{NaN}})\vee \bigvee_{k\in [0,m]} (\mathsf{epsAfter(k)} \wedge \mathsf{toNumber(k)} )$$
%It says either $x$ is not a number and so $n=-1$ or the first $k$ variables represent the value of $n$, for some $k\in [0,m]$. 
%$$
%\underf R {\psi_s} \defeq \Psi_n \wedge \Psi_{v_0}\wedge \parikhfof {R(x)} 
%$$
%To formulate the correctness lemma similar to Lemmas~\ref{lemma:memcorrect} and \ref{lemma:eqcorrect}, we need to consider a modified semantics of $R(x)$.
%Namely, we assume that $\semof{R(x)}$ is returns $\decode A(\parikhof{R(x)} \land )$

%We add $\Psi_n$ to $\Phi_O$.
%
%
%
%The correctness of this procedure is expressed in Lemma~\ref{lemma:sticorrect} below. We separate the lemmas to cases where $n\neq -1$ and $n=-1$. 
%
%
%\newcommand\restrict[1]{\raisebox{-.5ex}{$\big|$}_{#1}}

%\begin{lemma}\label{lemma:sticorrect}
%	
%	The interpretation $(I\cup\pim \cup I_n)  \in \semof {\Psi_n \wedge \Psi_{v_0}\wedge \parikhfof {A_x} }_{V_x \cup \#V_x \cup \{n\}} $ if and only if 
%	\begin{center}
%	 $\left(\begin{array}{c} 
%	 I_n(n)= \sti{\decode {A_x}(I\cup\pim)} \ \wedge \\
%	 I(v_i)=-1 \wedge i<j \Longrightarrow  I(v_j)=-1
%	 \end{array}\right)$\\ or\\
%	 $\left(\begin{array}{c} 
%	 I_n(n)= -1\ \wedge \decode {A_x}(I\cup\pim) \not\in [0,9]^+
%	 \end{array}\right)$
%	\end{center}
%
%\end{lemma}
%\yfc{sorry still ugly, the shifting of epsilon makes it a mess here}
%(An alternative description)
%\begin{lemma}\label{lemma:sticorrect}
%	For all $w \in \semof{A_{x}}$ and $c \in \mathbb{Z}$, we have 
%	\begin{center}
%		$\exists (I\cup\pim \cup I_n)  \in \semof {\Psi_n \wedge \Psi_{v_0}\wedge \parikhfof {A_x} }_{V_x \cup \#V_x \cup \{n\}}$ s.t. $w=\decode {A_x}(I\cup\pim)$ and $I_n(n)=c$\\
%		iff\\
%		$c= \sti{w}$.
%	\end{center}
%	
%\end{lemma}
%\hide{
%\begin{lemma}\label{lemma:sticorrect_pos}
%	$\semof {\Psi_n \wedge \Psi_{v_0}\wedge \parikhfof {A_x} }_{V_x \cup \#V_x \cup \{n\}} = \{(I\cup\pim \cup I_n) \mid  I_n(n)= \sti{\decode {A_x}(I\cup\pim)}\} $. 
%\end{lemma}
%In this case, $x$ encodes a number. The lemma says when the models we add to $\Phi_O$ represents a word in $\semof{A_x}$ whose integer value is exactly the model for $n$.
%
%\begin{lemma}\label{lemma:sticorrect_neg}
%	$\semof {\Psi_n \wedge \parikhfof {A_x} \wedge (v_0 = 0)\wedge (n = -1) }_{V_x \cup \#V_x \cup \{n\}} = 
%	\{(I\cup\pim \cup I_n) \mid  I_n(n)= -1 \wedge \decode {A_x}(I\cup\pim) \not\in [0,9]^+  \}$
%\end{lemma}
%
%In this case, $x$ does not encode a number. The lemma says  the models we add to $\Phi_O$ represents words in both $\semof{A_x}$ and $[0,9]^+$, which means not a number, and $n$ is $-1$.}
%
%\lh{
%\begin{lemma}
%$
%\semof {\Psi_n \wedge \Psi_{v_0}\wedge \parikhfof {A_x} }_{V_x \cup \#V_x \cup \{n\}} = \encode  R(\semof{\phi_s})
%$
%\end{lemma}
%The $n$ must somehow be handled properly by $\encode  R(\phi_s)$, I did not have time to think about it yet.
%}

% \section{Evaluation}
% \label{section:evaluation}
% \todo{copy a few tables from our test site to here}

@@

The most relevant work to ours is~\cite{abdulla2018trau} who proposed to project the solution space of variables to a generalization of flat automata. The main difference is that our approach works fully symbolically, which is enabled by to the ideas of using character variables and hence PFAs. The use of character variables allows our approach to handle string-number constraints efficiently -- the values and number of occurrences of those variables can be directly convert to a number in a linear formula. Moreover, the approach of~\cite{abdulla2018trau} is exponential to the alphabet size (we call this the alphabet-explosion problem in the introduction), and hence in the implementation, usually requires heuristics to prune out unnecessary symbols.

@@

\section{Conclusion and Future Works}
\label{section:conclusion}
In this paper, we report a novel approach for solving string constraints with string-number conversion and implemented it as an open-source tool {\tool}. For now, it support basic string constraints, string-number conversion, and also operations that can be encoded to them (e.g., \textsc{contains}, \textsc{prefixOf}). Since {\tool} is built inside the SMT solver Z3, we also get the power of processing formulae in the combination of different theories (e.g., array). Hence our tool can support the encoding of a wide range of program expressions. There are several avenues for future works. First, we are planning to integrate it with the JavaScript symbolic executor \textsc{cosette}~\cite{santos2018symbolic}. We  believe such integration  is feasible. 
%have discussed this possibility of such integration with some of its team members, and 
We are also planning to merge  {\tool}  with the main branch of the Z3 solver. For technical development, we think it would be interesting to consider the (symbolic) flattening of an even larger set of string operations, such as the one containing  \textsc{replaceAll} and \textsc{split}.

\begin{acks}
This research was  partially supported by the  Swedish Foundation for Strategic Research.
\end{acks}

\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

>>>>

\usepackage{balance}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\hide}[1]{}
\newcommand{\tool}{{\textsf{Z3-Trau}}}
\newcommand{\dr}{\mathcal{R}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\integers}{\mathbb{Z}}
\newcommand{\todo}[1]{{\color{blue}TODO: #1}}
\newcommand{\lh}[1]{{\color{orange}Lukas: #1}}
\newcommand{\changed}[1]{{\color{blue}#1}}

@@

However, the formula is satisfiable when $x=``00010"$. Observe that $\sti{``00010"}=10$ \changed{since PFAs accept  numeric words with least significant bit first}. The key is that even for a bounded integer, the corresponding numeral can be of an unbounded length with arbitrarily many trailing `$0$'s at the front. All numbers with up to $m$-digits can be however still handled without having to solve exponential constraints. It is enough to equip the initial state of the PFA with a  $0$-self-loop.

@@

	\caption{PFA for string-number conversion constraints}
	\label{fig:sfa_its}
	\Description{Parametric flat automata  for string-number conversion constraints}
\end{figure}

Numeric PFA with these restrictions would satisfy our primary objective, that is, they would induce linear formulae and would ``easily" and completely cover all numerals.
%
A last problem still needs to be solved before they can be efficient in practice. 
%
Recall that the character variables can be assigned $\epsilon$.  
%
Therefore, a single chain of $k$ interesting digits, $k\leq m$, can be by $A^m$ represented in $k\choose m$ ways, each corresponding to one possible interleaving of $k$ digits with $m-k$ epsilons. 
%
%\lh{I am not sure whether it is really impossible to make a small formula}
%
This may  lead to a formula of   exponential size  when defining  the value of $n$.
%
In order to eliminate this potential blow-up in the size of the formula, we add to $\paf^m$ an additional constraint that forces all epsilons to be \emph{shifted} behind the least significant digit. This will leave us with only one interleaving. This is the formula
$$
\Psi_{\mathsf{shift}}^m \defeq \bigwedge_{1\leq i \leq m} v_i \neq \noepsilon \implies v_{i-1}\neq \noepsilon \ .
$$
Last, since this restriction is meaningful only when the string is indeed a numeral,  
we also define the constraint representing the strings which are not numerals, the formula
$$
\Psi_{\mathsf{NaN}}^m \defeq  \bigvee_{i\in [1,m]} v_i>9
$$ 
and define the final form of the interpretation restriction used by $A^m$ as 
$$
\paf^m \defeq  \Psi_{\mathsf{NaN}}^m \lor (\Psi_{v_0} \land \Psi_{\mathsf{shift}}^m)\ .
$$
Consequently, we design our domain restrictions $\dr$ so that for string variables $x$ that appear within string-integer constraints,  $\dr(x)$ is a number PFA $(A^m,\paf^m),m\in \nat$.

@@

@@

\changed{
The most relevant work to ours is the FA-based approach~\cite{abdulla2018trau} that projects the solution space of variables to a generalization of flat automata. The main difference is that our approach works fully symbolically, which is enabled by to the ideas of using integer variables as characters and hence PFAs. The use of integer variables as characters allows our approach to handle string-number constraints efficiently -- the values and number of occurrences of those variables can be directly convert to a number in a linear formula.

The straightforward extension of existing FA based techniques boils down to solving the satisfiability problem of integer polynomial-exponential constraints, whose decidability is still open (Section~\ref{section:s2i}).

In principle, we can also extend FA-based solvers to support string-number conversion, but PFA are much more efficient. Extending FA-based approach to support string-number conversion would require multiple if-then-else statements (e.g., saying that if a transition labeled `0' is taken, then the corresponding number is $0$), which introduces a significant amount of additional predicates that the DPLL engine needs to decide their value. We have confirmed this in our preliminary experiment.}

\changed{
PASS\cite{goudong2013pass} uses quantified formulae over arrays with symbolic length to encode string constraints, and a specialized quantifier elimination to solve them. Though it differs from our approach significantly and handling the quantification is expensive, PASS is indeed similar in how finite automata and string-number conversion constraints are translated to formulae. But our translation is in general more efficient because it uses only linear arithmetics. In contrast, PASS translates string constraints to formulae with quantifier and array predicates. We did not compare with PASS empirically. The authors informed us that PASS is no longer maintained and is not open-source.
}

@@

\section{Conclusion and Future Works}
\label{section:conclusion}
In this paper, we report a novel approach for solving string constraints with string-number conversion and implemented it as an open-source tool {\tool}. For now, it support basic string constraints, string-number conversion, and also operations that can be encoded to them (e.g., \textsc{contains}, \textsc{prefixOf}). Since {\tool} is built inside the SMT solver Z3, we also get the power of processing formulae in the combination of different theories (e.g., array). Hence our tool can support the encoding of a wide range of program expressions. There are several avenues for future works. First, we are planning to integrate it with the JavaScript symbolic executor \textsc{cosette}~\cite{santos2018symbolic}. We  believe such integration  is feasible. 
We are also planning to merge  {\tool}  with the main branch of the Z3 solver. For technical development, we think it would be interesting to consider the (symbolic) flattening of an even larger set of string operations, such as the one containing  \textsc{replaceAll} and \textsc{split}.

\begin{acks}
This research was  partially supported by the  Swedish Foundation for Strategic Research and Ministry of Science and Technology Project,Taiwan (no. 106-2221-E-001 -009 -MY3).
\end{acks}

 \balance
\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}
